---
Author: Vaibhav Pitliya
College: Veermata Jijabai Technological Institute
Email: vaibhavpitliya28@gmail.com
Title: Analyzing factors causing delay and cancellation
output:
  html_document:
    df_print: paged
---
##1.Loading Library
```{r comment = NA, message = FALSE, warning = FALSE}
library(psych)
library(lattice)
library(corrplot)
library(corrgram)
library(car)
library(arm)
```
##2.Reading Data 
```{r comment = NA, message = FALSE, warning = FALSE}
setwd("C:/Users/vaibhav/Desktop/DataSets")
flights.df<-read.csv("flights.csv")
```
```{r comment = NA}
setwd("C:/Users/vaibhav/Desktop/DataSets")
airlines.df<-read.csv("airlines.csv")
airport.df<-read.csv("airports.csv")
```

##3.Cleaning Data
###Removing unwanted Coloumns
```{r comment = NA, message = FALSE, warning = FALSE}
flights_new.df<-cbind.data.frame(flights.df$YEAR,flights.df$MONTH,flights.df$DAY,flights.df$DAY_OF_WEEK,flights.df$AIRLINE,flights.df$FLIGHT_NUMBER,flights.df$ORIGIN_AIRPORT,flights.df$DESTINATION_AIRPORT,flights.df$DEPARTURE_DELAY,flights.df$AIR_TIME,flights.df$DISTANCE,flights.df$ARRIVAL_DELAY,flights.df$DIVERTED,flights.df$CANCELLED,flights.df$CANCELLATION_REASON)
flights_new.df$`flights.df$CANCELLATION_REASON`=factor(flights_new.df$`flights.df$CANCELLATION_REASON`,levels=c('A','B','C','D'),labels=c(1,2,3,4))

```


##4Finding Reasons for Cancellation

###Finding % of cancelled flights
```{r  comment = NA, message = FALSE, warning = FALSE}
cancelled_flights.df<-subset(flights_new.df,flights_new.df$`flights.df$CANCELLED`==1)
nrow(cancelled_flights.df)/nrow(flights_new.df)*100
```
Percentage of Flights Cancelled:1.57
###Finding which reasons lead to maximum cancellations
```{r comment = NA, message = FALSE, warning = FALSE}
mytable1<-with(cancelled_flights.df,table(cancelled_flights.df$`flights.df$CANCELLATION_REASON`))
prop.table(mytable1)*100
barchart(cancelled_flights.df$`flights.df$CANCELLATION_REASON`,data=cancelled_flights.df,col="black")
```










28% of cancellations were due to Airline or carrier reasons
54% of cancellations were due to Weather reasons
17% of cancellations were due to National Air System reasons
0.02% of cancellations were due to Security Reasons
Hence maximum 54% of cancellations are due to weather reasons

###Analyzing which month got maximum cancellations
```{r comment = NA, message = FALSE, warning = FALSE}
mytable2<-with(cancelled_flights.df,table(cancelled_flights.df$`flights.df$MONTH`))
prop.table(mytable2)*100
hist( cancelled_flights.df$`flights.df$MONTH`,
        xlab="Count",
        ylab="Month",
        xlim=c(0,12),
        main="Cancellations per month",
        col=c("black")      
        )

```








Top 5 Months with maximum cancellations are
1)february
2)January
3)March
4)June
5)December
It can be clearly seen maximum Cancellations are in the season of winter due to heavy snowfall
This can be seen as shown by the following table:
```{r comment = NA, message = FALSE, warning = FALSE}
mytable3<-table(cancelled_flights.df$`flights.df$CANCELLATION_REASON`,cancelled_flights.df$`flights.df$MONTH`)
prop.table(mytable3,2)*100
plot(cancelled_flights.df$`flights.df$CANCELLATION_REASON`,cancelled_flights.df$`flights.df$MONTH`)


```

###Analyzing Which Airlines got maximum cancellations
```{r message = FALSE, warning = FALSE}
mytable4<-with(cancelled_flights.df,table(cancelled_flights.df$`flights.df$AIRLINE`))
prop.table(mytable4)*100
plot(cancelled_flights.df$`flights.df$AIRLINE`,
     main="Maximum cancellations by which airlines",
     col="black")
```
Top 5 Airlines with maximum cancellations are
1)WN :Southwest Airlines Co.
2)EV :Atlantic Southeast Airlines
3)MQ :American Eagle Airlines Inc.
4)AA :American Airlines Inc.
5)OO : Skywest Airlines Inc.

###Analyzing  on which day flights got maximum cancellations
```{r comment = NA, message = FALSE, warning = FALSE}
mytable5<-with(cancelled_flights.df,table(cancelled_flights.df$`flights.df$DAY_OF_WEEK`))
prop.table(mytable5)*100
hist(cancelled_flights.df$`flights.df$DAY_OF_WEEK`,
     main="Cancellation of flights based on which day of week",
     col=c("black"))
```







In decreasing order of cancellations
1)Monday
2)Tuesday
3)Sunday
4)Thursday
5)Wednesday
6)Friday
7)Saturday


###Correlation between independent variables
```{r comments=NA, message = FALSE, warning = FALSE}
corr.test(cancelled_flights.df[,c(2,3,4,6,11,14)])
corrplot.mixed(cor(cancelled_flights.df[,c(2,3,4,6,11,14)],use="complete.obs"),upper="circle",tl.pos = "lt")
```

##5 DELAY RATIOS

###Distributing data in terms of few factors
Creating a new coloumn which states the factors about delays and cancellation
```{r message = FALSE, warning = FALSE}
flights_new.df$status<-'ON TIME'
flights_new.df$status[flights_new.df$`flights.df$DEPARTURE_DELAY`>0] <-'DEPARTURE DELAY'
flights_new.df$status[flights_new.df$`flights.df$CANCELLED`==1] <-'CANCELLED FLIGHTS'
flights_new.df$status[flights_new.df$`flights.df$ARRIVAL_DELAY`>0 & (flights_new.df$`flights.df$ARRIVAL_DELAY`-flights_new.df$`flights.df$DEPARTURE_DELAY`)>0 & flights_new.df$`flights.df$DEPARTURE_DELAY`>0] <-'DEPR & ARVL_DELAY'
flights_new.df$status[flights_new.df$`flights.df$ARRIVAL_DELAY`>0 & flights_new.df$`flights.df$DEPARTURE_DELAY`<=0] <-'ARRIVAL_DELAY'
flights_new.df$status<-factor(flights_new.df$status)
```
###Seeing which delays occured the most
```{r, message = FALSE, warning = FALSE}
mytable4<-table(flights_new.df$status)
prop.table(mytable4)*100
plot(flights_new.df$status)
```






52% flights were on time
26% flights had departure delays
9% flights had arrival delays
9% flights had arrival and departure delay
1.54% flights were cancelled


##6 DEPARTURE DELAY ANALYSIS
###Summarizing departure delay
```{r message = FALSE, warning = FALSE}
summary(flights_new.df$`flights.df$DEPARTURE_DELAY`)
```
###Breaking departure delay as per the time
```{r message = FALSE, warning = FALSE}
flightsDepDelay<-subset(flights_new.df,flights_new.df$`flights.df$DEPARTURE_DELAY`>0)
flightsDepDelay$DelayTimeInterval<-cut(flightsDepDelay$`flights.df$DEPARTURE_DELAY`,
                                        breaks= c(0,10,30,60,90,120,180,1988),
                                        labels=c("(0-10]","(10-30]","(30-60]","(60-90]","(90-120]","(120-180]","(180-1988]"))
```


###Tabluating departure delays
```{r message = FALSE, warning = FALSE}
mytable5<-with(flights_new.df,table(flightsDepDelay$DelayTimeInterval))
prop.table(mytable5)*100
plot(flightsDepDelay$DelayTimeInterval,col="Black")
```





Hence
Of all the Departure delays
41% flights were delayed by less tha 10 mins
28% flights were delayed by 10 to 30 mins
14% flights were delayed by 30 to 60 mins
6% flights were delayed by 60 to 90  mins
3% flights were delayed by 90 to 120 mins
3% flights were delayed by 120 to 180 mins
2% flights were delayed by 180+ mins


```{r}
mytable6<-table(flightsDepDelay$DelayTimeInterval,flightsDepDelay$`flights.df$MONTH`)
prop.table(mytable6,1)*100
```
### Running Correlation tests
```{r}
corrplot.mixed(cor(flightsDepDelay[,c(2,3,4,6,9,11,13)]),upper="ellipse",tl.pos="lt")
```
 


### Distributing data in terms of 3 factors for easier evaluation
```{r}
flightsDepDelay$Latency[flightsDepDelay$`flights.df$DEPARTURE_DELAY`>0 & flightsDepDelay$`flights.df$DEPARTURE_DELAY`<5]<-'On time'
flightsDepDelay$Latency[flightsDepDelay$`flights.df$DEPARTURE_DELAY`>=5 & flightsDepDelay$`flights.df$DEPARTURE_DELAY`<15]<-'Small Delay'
flightsDepDelay$Latency[flightsDepDelay$`flights.df$DEPARTURE_DELAY`>=15]<-"Long Delay"
flightsDepDelay$Latency<-factor(flightsDepDelay$Latency)

```

```{r}
mytable7<-with(flightsDepDelay,table(flightsDepDelay$Latency))
prop.table(mytable7)*100
```
###Seeing these 3 factors with respect to month
```{r}
mytable8<-table(flightsDepDelay$`flights.df$MONTH`,flightsDepDelay$Latency)
prop.table(mytable8,2)*100
```





Of all the long delays 
Maximum were in the month of:
1)June
2)July
3)March,August,December


###Seeing these 3 factors with respect to day of week
```{r}
mytable9<-table(flightsDepDelay$`flights.df$DAY_OF_WEEK`,flightsDepDelay$Latency)
prop.table(mytable9,2)*100
```



All percentages are almost same so no evident conclusion can be drawn


###Seeing these 3 factors with respect to Airlines
```{r}
mytable10<-table(flightsDepDelay$`flights.df$AIRLINE`,flightsDepDelay$Latency)
prop.table(mytable10,2)*100
```

-Maximum Long delays are seen in Southwest Airlines Co.(25%) folowed by American Airlines Inc.,Delta Air Lines Inc.,United Air Lines Inc.
-Similar trend is seen in small delays and on time flights too.


###Some insights(Comparing average delay time of each Airline)
```{r}
var1<-aggregate(flightsDepDelay$`flights.df$DEPARTURE_DELAY`,list(flightsDepDelay$`flights.df$AIRLINE`),mean)
var1
plot(x=var1$Group.1,y=jitter(var1$x))
```




This shows that when taking into account no. of flights along with delay time maximum or the airlines with highest delay is F9 folowed by NK and EV

Hence Ailines affecting it changes drastically after considering number of flights

##7 ARRIVAL DELAYS
###Summarizing arrival delay
```{r}
summary(flights_new.df$`flights.df$ARRIVAL_DELAY`)
```
###Breaking arrival delay as per the time
```{r}
flightsArrDelay<-subset(flights_new.df,!is.na(flights_new.df$`flights.df$ARRIVAL_DELAY`))
flights_ArrDelay<-subset(flightsArrDelay,flightsArrDelay$`flights.df$ARRIVAL_DELAY`>0)
flights_ArrDelay$DelayTimeInterval<-cut(flights_ArrDelay$`flights.df$ARRIVAL_DELAY`,
                                        breaks= c(0,10,30,60,90,120,180,1988),
                                        labels=c("(0-10]","(10-30]","(30-60]","(60-90]","(90-120]","(120-180]","(180-1988]"))
```

###Tabulating Arrival Delay
```{r}
mytable11<-with(flights_new.df,table(flights_ArrDelay$DelayTimeInterval))
prop.table(mytable11)*100
plot(flights_ArrDelay$DelayTimeInterval,col="Black")
```


Hence
Of all the Arrival delays
39% flights were delayed by less tha 10 mins
29% flights were delayed by 10 to 30 mins
15% flights were delayed by 30 to 60 mins
6% flights were delayed by 60 to 90  mins
3% flights were delayed by 90 to 120 mins
3% flights were delayed by 120 to 180 mins
2% flights were delayed by 180+ mins

This is very similar to Departure delays distribution

### Running Correlation tests
```{r}
corrplot.mixed(cor(flights_ArrDelay[,c(2,3,4,6,9,12,13)]),upper="ellipse",tl.pos="lt")
```
 







### Distributing data in terms of 3 factors for easier evaluation
```{r}
flights_ArrDelay$Latency[flights_ArrDelay$`flights.df$ARRIVAL_DELAY`>0 & flights_ArrDelay$`flights.df$ARRIVAL_DELAY`<5]<-'On time'
flights_ArrDelay$Latency[flights_ArrDelay$`flights.df$ARRIVAL_DELAY`>=5 & flights_ArrDelay$`flights.df$ARRIVAL_DELAY`<15]<-'Small Delay'
flights_ArrDelay$Latency[flights_ArrDelay$`flights.df$ARRIVAL_DELAY`>=15]<-"Long Delay"
flights_ArrDelay$Latency<-factor(flights_ArrDelay$Latency)

```

```{r}
mytable12<-with(flights_ArrDelay,table(flights_ArrDelay$Latency))
prop.table(mytable12)*100
```



50% Flights have long delay
30% Flights have small delay

###Seeing these 3 factors with respect to month
```{r}
mytable13<-table(flights_ArrDelay$`flights.df$MONTH`,flights_ArrDelay$Latency)
prop.table(mytable13,2)*100
```





Of all the long delays 
Maximum were in the month of:
1)June
2)July
3)Jan,December


###Seeing these 3 factors with respect to day of week
```{r}
mytable14<-table(flights_ArrDelay$`flights.df$DAY_OF_WEEK`,flights_ArrDelay$Latency)
prop.table(mytable14,2)*100
```



All percentages are almost same so no evident conclusion can be drawn


###Seeing these 3 factors with respect to Airlines
```{r}
mytable15<-table(flights_ArrDelay$`flights.df$AIRLINE`,flights_ArrDelay$Latency)
prop.table(mytable15,2)*100
```

-Maximum Long delays are seen in Southwest Airlines Co.(25%) folowed by American Airlines Inc.,Delta Air Lines Inc.,United Air Lines Inc.
-Similar trend is seen in small delays and on time flights too.


###Some insights(Comparing average delay time of each Airline)
```{r}
var3<-aggregate(flights_ArrDelay$`flights.df$ARRIVAL_DELAY`,list(flights_ArrDelay$`flights.df$AIRLINE`),mean)
var3
plot(x=var3$Group.1,y=jitter(var3$x))
```




This shows that when taking into account no. of flights along with delay time maximum or the airlines with highest delay is F9 folowed by NK,UA and MQ

## 8 REGRESSION MODELS
###Running a correlation test
```{r}
cor.test(flights_new.df$`flights.df$ARRIVAL_DELAY`,flights_new.df$`flights.df$DEPARTURE_DELAY`)
```


###Predicting Departure delay based on various factors
```{r}
fit1<-lm(flights.df$DEPARTURE_DELAY~flights.df$AIR_SYSTEM_DELAY+flights.df$WEATHER_DELAY+flights.df$LATE_AIRCRAFT_DELAY+flights.df$AIRLINE_DELAY+flights.df$SECURITY_DELAY)
summary(fit1)
```

Hence we can predict departure delay based on other delay
This model is quite good at predicting delays as Adjusted R-squared is 0.9402

###Predicting Arrival delay based on various factors
```{r}
fit2<-lm(flights.df$ARRIVAL_DELAY~flights.df$AIR_SYSTEM_DELAY+flights.df$WEATHER_DELAY+flights.df$LATE_AIRCRAFT_DELAY+flights.df$AIRLINE_DELAY+flights.df$SECURITY_DELAY)
summary(fit2)
```

###Predicting Arrival delay based on Departure delay
```{r}
fit<-lm(flights_new.df$`flights.df$ARRIVAL_DELAY`~flights_new.df$`flights.df$DEPARTURE_DELAY`)
summary(fit)
```
Hence Arrival Delay can be predicted with Departure Delay as multiple R squared = 0.8924 


